---
title: "Appendix 2: Lake Washington Analysis"
format: pdf
editor: visual
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

# Appendix 2: Lake Washington Analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,  
  error = FALSE,    
  message = FALSE
)
```

```{r, include=FALSE}
library(MARSS)
library(mgcv)
library(dplyr)
library(forecast)
library(ggplot2)
library(lubridate)
library(viridis)
library(paletteer)
library(ggpubr)
library(patchwork)
col<-paletteer_d("nationalparkcolors::Everglades")
```

Here we show the GAM and DLM analysis for the Lake Washington dataset. For complete code, see the [full quarto document](https://github.com/ecosystem-state/nonstationarity-best-practices/blob/main/Lake_WA_case_study/Appendix2_LakeWashington.qmd). We include some code snippets here for reference.

## Data Processing

```{r, include=FALSE}
dat <- as.data.frame(MARSS::lakeWAplanktonRaw)
dat$date <- make_date(dat$Year, dat$Month, day = 1)
```

In Lake WA the major change ecosystem change is related to sewage in the lake – adding wastewater plants in the 1960s reduced phosphorous (@suppfig-TSphos) and in turn, blue green algae (@suppfig-TSbg) which can be seen by examining time series plots of the data.

```{r, include=FALSE}
tsphos <- ggplot(dat, aes(date, TP)) + 
  geom_point() + ylab("Total phosphorus") + xlab("") + theme_classic()
tsbg <- ggplot(dat, aes(date, log(Bluegreens+1))) + 
  geom_point() + ylab("Ln Blue-green algae") + xlab("") + theme_classic()
#gridExtra::grid.arrange(p1,p2,nrow=2)

```

We can then use the MARSS package to fit a DLM. We'll filter the data to only use counts before 1974, and log transform densities. We also z-score the covariate. This is a good example of imperfect data – there's a couple of missing observations in the covariate, which we can't have in a DLM or GAM – so let's just fill them in with a spline. We can do this as follows:

```{r}
dat <- dplyr::filter(dat, Year < 1974) %>%
  dplyr::mutate(y = log(Bluegreens), zp = as.numeric(scale(TP)))

interpolated <- spline(dat$zp, xout = 1:nrow(dat))$y
dat$zp[which(is.na(dat$zp))] <- interpolated[which(is.na(dat$zp))]

```

The response data exhibit a strong seasonal cycle – and it's a good idea to de-season that to focus on the phosphorous effect. Here we de-season the data:

```{r}
# Fill in a few missing values
dat$y_interp <- dat$y
interpolated <- spline(dat$y, xout = 1:nrow(dat))$y
missing <- which(is.na(dat$y_interp))
dat$y_interp[missing] <- interpolated[missing]

dat <- dplyr::group_by(dat, Month) %>%
  dplyr::mutate(month_mean = mean(y,na.rm=T)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(y_adj = y_interp - month_mean)

```

Last, we should look at the lags between driver and response relationships which we can do using a CCF (@suppfig-CCF). The CCF here indicates a 2-4 month lag, where TP and algae are more strongly associated with one another.

We'll use a lag of 4 months, which we can apply as follows:

```{r}

lag_n <- 5
dat$lagged_zp <- c(rep(NA, lag_n), dat$zp[1:(nrow(dat)-lag_n)]) 
dat <- dat[-c(1:lag_n),]
```

## DLMs

First, we need to define the model – this block is basically copied from the MARSS book, [the salmon survival case study](https://atsa-es.github.io/atsa-labs/sec-dlm-fitting-a-univariate-dlm-with-marss.html)

```{r}
m <- 2
TT <- nrow(dat)
B <- diag(m) ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1) ## 2x1; both elements = 0
Q <- matrix(list(0), m, m) ## 2x2; all 0 for now
diag(Q)[1] <- 0.0000001
diag(Q)[2] <- c("q.beta")
#diag(Q) <- c("q.alpha", "q.beta") ## 2x2; diag = (q1,q2)
Z <- array(NA, c(1, m, TT)) ## NxMxT; empty for now
Z[1, 1, ] <- rep(1, TT) ## Nx1; 1's for intercept
Z[1, 2, ] <- dat$lagged_zp ## Nx1; predictor variable
A <- matrix(0) ## 1x1; scalar = 0
R <- matrix("r") ## 1x1; scalar = r
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))
## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
# convert response to matrix
dat_mat <- matrix(dat$y_adj, nrow = 1)
# fit the model -- crank up the maxit to ensure convergence
dlm_1 <- MARSS(dat_mat, inits = inits_list, model = mod_list,
control = list(maxit=4000), method="TMB")



```

Let’s put the state estimates back into the original dataframe and do some plotting. The slope is generally positive, which is what we expect for the hypothesized relationship between bluegreen algae and total phosphorus – and exhibits clear variation through time (@suppfig-dlmSlope). We can also plot the fitted values from our model (@suppfig-dlmFit).

```{r, echo=FALSE}
dat$int_est <- dlm_1$states[1,]
dat$int_se <- dlm_1$states.se[1,]

dat$slope_est <- dlm_1$states[2,]
dat$slope_se <- dlm_1$states.se[2,]
DLM_plot <- ggplot(dat, aes(date, slope_est)) +
geom_ribbon(aes(ymin=slope_est-2*slope_se, ymax = slope_est+2*slope_se), alpha=0.6) +
geom_line() + ylab("Time-varying slope") + xlab("") + theme_classic()
```

```{r, echo=FALSE}
pred <- fitted(dlm_1)
dat$pred <- pred$.fitted
dlmFit<-ggplot(dat, aes(date, pred)) +
geom_line() +
ylab("De-seasoned Ln blue-green algae") +
xlab("Date") +
theme_classic() +
geom_point(aes(date,y_adj), col="red",alpha=0.5)
```

And we can examine the residuals for a temporal trend, and see that because the DLM is flexible, there's not much evidence of a trend (@suppfig-dlmresid) and there are no issues wiht the q-q plot (@suppfig-dlmqq).

```{r}
#| echo: false
#| include: false
#| error: false
#| warning: false
resid_dlm <- predict(dlm_1)$pred
resid_dlm$resid <- resid_dlm$y - resid_dlm$estimate
resid_dlm$date <- dat$date

dlm_resid<- ggplot(resid_dlm, aes(date, resid)) + 
  geom_point() + 
  geom_smooth() + 
  theme_classic() + xlab("Date") + ylab("Residuals") + ggtitle("DLM Residuals")
```

```{r}
#| echo: false
#| include: false
#| error: false
#| warning: false
dlm_qq<-ggplot(resid_dlm, aes(sample = resid)) + stat_qq() + stat_qq_line() + labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) + theme_classic() + ggtitle("DLM Residuals")
```

## GAMs

We can fit a GAM model 5 ways, 1) so that the smoother is a function of phosphorus which represents a nonlinear functional relationship between total phosphorus and bluegreen algae that is not temporally structured thus the relationship is stationary and fixed through time. This is how GAMs are typically structured for ecological analyses. 2) a temporally structured model such that the abundance of bluegreen algae varies as a function of time unrelated to phsophorus, 3) A temporally structured model so that the smooth function of phosphorus varies by time, and 4) a smooth on date as a driver of cyanobacteria where the smooth through time varies by phosphorus level. Finally, these relationships can be fit as 5) a tensor product which provide a multivariate function spanning the dimensions of time and phosphorus. We will not discuss tensor products in greater detail and only include them for completeness. We show the model predicted cyanobacteria trend through time using these five modeling approaches (@suppfig-gamFit). While they are quite similar, the predictions diverge sustantially at the end of the time series.

```{r}

dat$n_date <- as.numeric(as.factor(dat$date))

gam_cr1 <- gam(y_adj ~ s(lagged_zp, bs = "cr"), data = dat) #1
gam_cr2 <- gam(y_adj ~ s(lagged_zp, by = n_date, bs = "cr"), data = dat) #2
gam_cr3 <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr"), data = dat) #3
gam_cr4 <- gam(y_adj ~ s(n_date), data = dat) #4
gam_cr5 <- gam(y_adj ~ te(lagged_zp, n_date), data = dat) #5
```

```{r, include=FALSE}
plot_cr1 <- plot(gam_cr1, seWithMean = TRUE, n = nrow(dat))
df_cr1 <- data.frame(x = plot_cr1[[1]]$x,
est = plot_cr1[[1]]$fit,
se = plot_cr1[[1]]$se)
```

```{r, echo=FALSE}
p1 <- ggplot(df_cr1, aes(x, est)) +
geom_ribbon(aes(ymin = est - 2*se, ymax = est + 2*se), alpha = 0.4, col = NA, fill=viridis(1)) +
theme_classic() +
geom_line(col = viridis(1)) + 
xlab("Phosphorous") +
ylab("Effect")
#ggsave(p1, filename = "Figure_S1_nonlinear_smooth.png", height = 5, width=7)
# make second plot showing fits from all 3 models

# Predict for all models and store model label (corresponding to main fig)
pred_1 <- dat
pred_1$pred <- predict(gam_cr1, se.fit = TRUE)$fit
pred_1$se <- predict(gam_cr1, se.fit = TRUE)$se.fit
pred_1$smooth <- "Panel C: s(Phos)"

pred_2 <- dat
pred_2$pred <- predict(gam_cr2, se.fit = TRUE)$fit
pred_2$se <- predict(gam_cr2, se.fit = TRUE)$se.fit
pred_2$smooth <- "Panel E: s(Phos, by=date)"

pred_3 <- dat
pred_3$pred <- predict(gam_cr3, se.fit = TRUE)$fit
pred_3$se <- predict(gam_cr3, se.fit = TRUE)$se.fit
pred_3$smooth <- "Panel D: s(date, by=Phos)"

pred_4 <- dat
pred_4$pred <- predict(gam_cr4, se.fit = TRUE)$fit
pred_4$se <- predict(gam_cr4, se.fit = TRUE)$se.fit
pred_4$smooth <- "Panel B: s(date)"

pred_5 <- dat
pred_5$pred <- predict(gam_cr5, se.fit = TRUE)$fit
pred_5$se <- predict(gam_cr5, se.fit = TRUE)$se.fit
pred_5$smooth <- "Panel F: te(Phos, date)"

pred_6 <- dat
pred_6$se <- predict(gam_cr5, se.fit = TRUE)$se.fit
pred_6$smooth <- "DLM"

# Combine all predictions
d_all <- rbind(pred_4, pred_1, pred_3, pred_2, pred_5, pred_6)
d_allv2 <- rbind(pred_4, pred_1, pred_3, pred_2, pred_6)
pred<-pred%>%mutate(smooth="DLM")%>%
  mutate(date=as.Date(t))
# Create the plot (Panel A)
pA <- ggplot(d_all, aes(date, y_adj, col = smooth)) + 
  geom_point(col = "black", alpha = 0.4) + 
  geom_line(aes(y = pred), linewidth = 1.5, alpha = 0.7) + 
  theme_classic() + 
  xlab("") + 
  ylab("Deseasoned cyanobacteria") + 
  scale_color_viridis_d(
    option = "mako",
    begin = 0.1,
    end = 0.9,
    name = "Model"
  ) +
  labs(title = "Observed vs. predicted fits from all GAMs")
  viridis(n = 5, option = "mako", begin = 0.1, end = 0.9)
d_allv2 <- rbind(pred_4, pred_1, pred_3, pred_2, pred_6)%>%
      mutate(smooth2 = factor(smooth, levels = c("DLM","Panel B: s(date)","Panel C: s(Phos)", "Panel D: s(date, by=Phos)","Panel E: s(Phos, by=date)")))
pAv2 <- ggplot(d_allv2, aes(date, y_adj, col = smooth2)) + 
  geom_point(col = "black", alpha = 0.4) + 
  geom_line(aes(y = pred), linewidth = 1.5, alpha = 0.7) + 
  theme_classic() + 
  xlab("Date") + 
  ylab("Seasonally detrended cyanobacteria") + 
  scale_color_manual(values=c( "#A8E1BCFF","#26172AFF", "#414081FF", "#357BA2FF", "#3DB4ADFF"),
    name = "Model"
  ) +
  labs(title = "Observed vs. predicted fits from all GAMs")

```

```{r, include= FALSE}
# Predict smooth over date
# Predict smooth over date
model_colors <- viridisLite::viridis(n = 5, option = "mako")
grid_B <- data.frame(n_date = dat$n_date, date = dat$date)
pred_B <- predict(gam_cr4, newdata = grid_B, se.fit = TRUE)
grid_B$fit <- pred_B$fit
grid_B$se <- pred_B$se

pB <- ggplot(grid_B, aes(date, fit)) +
  geom_line(color = model_colors[1], linewidth = 1.2) +
  geom_ribbon(aes(ymin = fit - 2*se, ymax = fit + 2*se), alpha = 0.3, fill = model_colors[1]) +
  theme_classic() +
  labs(
    title = "Smooth of Date: s(date)",
    x = "", y = "Effect"
  )

grid_C <- data.frame(lagged_zp = seq(min(dat$lagged_zp), max(dat$lagged_zp), length.out = 100))
pred_C <- predict(gam_cr1, newdata = grid_C, se.fit = TRUE)
grid_C$fit <- pred_C$fit
grid_C$se <- pred_C$se

pC <- ggplot(grid_C, aes(lagged_zp, fit)) +
  geom_line(color = model_colors[2], linewidth = 1.2) +
  geom_ribbon(aes(ymin = fit - 2*se, ymax = fit + 2*se), alpha = 0.3, fill = model_colors[2]) +
  theme_classic() +
  labs(
    title = "Smooth of Phosphorus: s(Phos)",
    x = "", y = "Effect"
  )

# This panel is for s(n_date, by = lagged_zp)
# We are going to use 3 reference values of phosphorous at the quantiles below
zp_vals <- quantile(dat$lagged_zp, probs = c(0.1, 0.5, 0.9))
grid_D <- expand.grid(
  lagged_zp = zp_vals,
  n_date = dat$n_date
)
grid_D$date <- dat$date[grid_D$n_date]
pred_D <- predict(gam_cr3, newdata = grid_D, se.fit = TRUE)
grid_D$fit <- pred_D$fit
grid_D$se <- pred_D$se
grid_D$group <- factor(grid_D$lagged_zp, labels = c("Low Phos", "Med Phos", "High Phos"))

pD <- ggplot(grid_D, aes(date, fit, color = group, fill = group)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = fit - 2*se, ymax = fit + 2*se), alpha = 0.2, color = NA) +
  scale_color_viridis_d(name = "Phosphorus", begin=0.2, end = 0.8) +
  scale_fill_viridis_d(name = "Phosphorus", begin=0.2, end = 0.8) +
  theme_classic() +
  labs(
    title = "Time effect by Phosphorus: s(date, by = Phos)",
    x = "Date", y = "Effect"
  )

# This panel is for s(lagged_zp, by = n_date)
grid_E <- expand.grid(
  lagged_zp = seq(min(dat$lagged_zp), max(dat$lagged_zp), length.out = 200),
  n_date = c(25, 73, 121)
)

pred_E <- predict(gam_cr2, newdata = grid_E, se.fit = TRUE)
grid_E$fit <- pred_E$fit
grid_E$se <- pred_E$se
grid_E$date_label <- factor(paste0("Day ", grid_E$n_date),
                            labels = c("June 1964", "June 1968", "June 1972"))

pE <- ggplot(grid_E, aes(lagged_zp, fit, color = date_label, fill = date_label)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = fit - 2 * se, ymax = fit + 2 * se), alpha = 0.2, color = NA) +
  scale_color_viridis_d(name = "Date", begin=0.2, end = 0.8) +
  scale_fill_viridis_d(name = "Date", begin=0.2, end = 0.8) +
  theme_classic() +
  labs(
    title = "Phosphorus effect by date: s(Phos, by = date)",
    x = "Z-scored Phosphorus", y = "Effect"
  )

grid_F <- expand.grid(
  lagged_zp = seq(min(dat$lagged_zp), max(dat$lagged_zp), length.out = 100),
  n_date = seq(min(dat$n_date), max(dat$n_date), length.out = 100)
)
pred_F <- predict(gam_cr5, newdata = grid_F, se.fit = TRUE)
grid_F$fit <- pred_F$fit

pF <- ggplot(grid_F, aes(n_date, lagged_zp, fill = fit)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(option = "C") +
  theme_classic() +
  labs(
    title = "Tensor interaction: te(Phos, date)",
    x = "Date", y = "Phosphorus",
    fill = "Effect"
  )
blank<-ggplot()+theme(plot.background = element_rect(fill = "white"))+ theme_void()
row2<-ggarrange(pB, blank, pC,blank,nrow=1,ncol=4,widths=c(4,1.3,4,1.3),labels=c("B.","","C.",""))
row3<-ggarrange(pD, pE,nrow=1,ncol=2,labels=c("D.","E."))
bottom_panels<-ggarrange(row2,row3,nrow=2,ncol=1)

final_plotv2 <- ggarrange(pAv2, bottom_panels,nrow=2,labels=c("A.", ""))
final_plot <- (pA + pB + pC + pD + pE) +
  plot_layout(ncol = 2, nrow = 3) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag.position = c(0.01, 0.98))

# Save the combined figure
ggsave(final_plotv2, filename="gam_comparison_gridv2.png", height = 8, width = 10, bg="white")
```

It is important to also examine the residuals from these models. For example, we can look at the residuals for the GAMs fit to the Lake Washington data. In looking across models, those that do not include time-varying parameters show evidence of a persistent trend in the residuals that is not accounted for by the model ((@suppfig-gamresid) despite teh q-q plot being reasonable for all model specifications ((@suppfig-gamqq).

```{r}

resid_df_1 <- data.frame(date = dat$date, resid = resid(gam_cr1),
                         Model = "s(Phos)")
resid_df_2 <- data.frame(date = dat$date, resid = resid(gam_cr2),
                         Model = "s(Phos, by = date)")
resid_df_3 <- data.frame(date = dat$date, resid = resid(gam_cr3),
                         Model = "s(date, by=Phos)")
resid_df_4 <- data.frame(date = dat$date, resid = resid(gam_cr4),
                         Model = "s(date)")
resid_df_5 <- data.frame(date = dat$date, resid = resid(gam_cr5),
                         Model = "te(Phos, date")
resid <- rbind(resid_df_1, resid_df_2, resid_df_3, resid_df_4, resid_df_5)
gam_resid<-ggplot(resid, aes(date, resid, group = Model)) + 
  geom_point() + 
  geom_smooth() + theme_classic() + 
  facet_wrap(~ Model) + xlab("Date") + 
  ylab("Residuals") + ggtitle("GAM Residuals")

```

```{r echo=FALSE}
gam_qq<- ggplot(resid, aes(sample = resid)) + stat_qq() + stat_qq_line() + labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) + theme_classic() + facet_wrap(~ Model) + ggtitle("GAM Residuals")
```

There are a few more analysis approaches for running GAMs including the choice of smoother function (@suppfig-gamFit & @suppfig-gamSmooth) and the number of knots, or the "wiggliness" (@suppfig-gamknots) of the smoothed relationship. We can plot these different results and include the results of the DLM for comparison, note that 'cr' and 'tp' are so similar they are difficult to discern from the plot. We can do a similar comparison for the choice of k. Here we compare a four different numbers of knots (@suppfig-gamknots), we show an intentionally wide range of k (30 - 120) to illustrate the effects of k on estimated uncertainty, in practice k should be specified based on degrees of freedom and expectations of variability through time.

```{r, echo=FALSE}

# The 'sp' parameter also controls the penalty on smoothing (below)
dat$n_date <- as.numeric(as.factor(dat$date))
gam_cr <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr",k=nrow(dat)), 
              data = dat)
gam_tp <- gam(y_adj ~ s(n_date, by = lagged_zp,k=nrow(dat)), 
              data = dat) 
gam_gp <- gam_cr <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "gp",k=nrow(dat)), 
                        data = dat)

```

```{r, include=FALSE}
dat$pred_gam <- predict(gam_tp)

gam_smooth<- ggplot(dat, aes(date, pred)) + 
  geom_line() + ylab("De-seasoned Ln blue-green algae") + 
  xlab("Date") + theme_classic() + 
  geom_point(aes(date,y_adj), col="red",alpha=0.5) + 
  geom_line(aes(date,pred_gam), col="blue",alpha=0.5)

# extract the smooth estimates with SEs to compare to DLM
plot_cr <- plot(gam_cr, seWithMean = TRUE, n = nrow(dat))
plot_tp <- plot(gam_tp, seWithMean = TRUE, n = nrow(dat))
plot_gp <- plot(gam_gp, seWithMean = TRUE, n = nrow(dat))
```

```{r, echo=FALSE}
df_cr <- data.frame(date = dat$date, 
                    est = plot_cr[[1]]$fit, 
                    se = plot_cr[[1]]$se, type="cr")
df_tp <- data.frame(date = dat$date, 
                    est = plot_tp[[1]]$fit, 
                    se = plot_tp[[1]]$se, type="tp")
df_gp <- data.frame(date = dat$date, 
                    est = plot_gp[[1]]$fit, 
                    se = plot_gp[[1]]$se, type="gp")
df_dlm <- data.frame(date = dat$date, 
                     est = dat$slope_est, 
                     se = dat$slope_se, type="dlm")
df_smooth <- rbind(df_cr, df_tp, df_gp, df_dlm)

#df_smooth <- rbind(df_cr,  df_gp)
df_smooth_type<-df_smooth
gam_smooth<-ggplot(df_smooth, aes(date, est, col=type, fill=type)) + 
  geom_line(aes(date,est-2*se), alpha=0.5, linetype=2) + 
  geom_line(aes(date,est+2*se), alpha=0.5, linetype=2) + 
  geom_line(aes(date,est), alpha=0.5) + 
  theme_classic() + 
  xlab("Date") + 
  ylab("Phosphorous effect on blue-green algae (Slope)") 
```

```{r, include=FALSE}
gam_cr30 <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr",
                          k=nrow(dat)), data = dat)
gam_cr60 <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr",
                          k=nrow(dat),sp=16), data = dat)
gam_cr90 <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr",
                          k=nrow(dat),sp=2.5), data = dat)
gam_cr120 <- gam(y_adj ~ s(n_date, by = lagged_zp, bs = "cr",
                           k=nrow(dat),sp=0.3), data = dat)

plot_cr30 <- plot(gam_cr30, seWithMean = TRUE, n = nrow(dat))
plot_cr60 <- plot(gam_cr60, seWithMean = TRUE, n = nrow(dat))
plot_cr90 <- plot(gam_cr90, seWithMean = TRUE, n = nrow(dat))
plot_cr120 <- plot(gam_cr120, seWithMean = TRUE, n = nrow(dat))

df_30 <- data.frame(date = dat$date, 
                    est = plot_cr30[[1]]$fit, 
                    se = plot_cr30[[1]]$se, knots="30")
df_60 <- data.frame(date = dat$date, 
                    est = plot_cr60[[1]]$fit, 
                    se = plot_cr60[[1]]$se, knots="60")
df_90 <- data.frame(date = dat$date, 
                    est = plot_cr90[[1]]$fit, 
                    se = plot_cr90[[1]]$se, knots="90")
df_120 <- data.frame(date = dat$date, 
                     est = plot_cr120[[1]]$fit, 
                     se = plot_cr120[[1]]$se, knots="120")
df_smooth <- rbind(df_30, df_60, df_90, df_120)
df_smooth_knots<-df_smooth
```

```{r, echo=FALSE}

gamknots <- ggplot(df_smooth, aes(date, est, col=knots, fill=knots)) + 
  geom_line(aes(date,est-2*se), alpha=0.5, linetype=2) + 
  geom_line(aes(date,est+2*se), alpha=0.5, linetype=2) + 
  geom_line(aes(date,est), alpha=0.5) + 
  theme_classic() + xlab("Date") + ylab("Phosphorous effect on blue-green algae (slope)") 

```

```{r, include=FALSE}

#Coding Figure 2 all together
# 
# panel.a <- ggplot(d_all, aes(date, y_adj, col = smooth)) + 
#   geom_point(col="grey") + 
#   theme_classic() + 
#   xlab("Year")+
#     geom_ribbon(aes(ymin = pred-2*se, ymax = pred+2*se, y = NULL, fill=smooth),
#               alpha = 0.15, colour = NA)+
#   scale_fill_manual(values=c(col[4],col[3],col[2]))+
#   scale_colour_manual(values=c(col[4],col[3],col[2]))+
#   geom_line(aes(date, pred), size=1.5,alpha=0.5) + 
#   xlab("") + ylab("Detrended cyanobacteria")  
#  # scale_color_viridis_d(option="magma",begin=0.2, end=0.8)
# 
# panel.b <-ggplot(df_smooth_knots, aes(date, est, col=knots, fill=knots)) + 
#   geom_line(aes(date,est-2*se), alpha=0.5, linetype=2) + 
#   geom_line(aes(date,est+2*se), alpha=0.5, linetype=2) + 
#      geom_ribbon(aes(ymin = est-2*se, ymax = est+2*se, y = NULL, fill=knots),
#               alpha = 0.1, colour = NA)+ 
#   geom_line(aes(date,est), alpha=0.6, lwd=0.75) + 
#   ylim(c(-17,21))+
#   scale_colour_manual(values=c(col[4],col[5],col[3],col[1]))+
#   scale_fill_manual(values=c(col[4],col[5],col[3],col[1]))+ 
#   #ggtitle("Effect of smoothing DF on SEs")+
#   theme_classic() + xlab("") + ylab("Phosphorous effect on blue-green algae \n (Time-varying slope)") 
# 
# panel.c<-ggplot(df_smooth_type, aes(date, est, col=type)) + 
#        geom_ribbon(aes(ymin = est-2*se, ymax = est+2*se, y = NULL, fill=type),
#               alpha = 0.05, colour = NA)+ 
#   geom_line(aes(date,est-2*se), alpha=0.5, linetype=2) + 
#   geom_line(aes(date,est+2*se), alpha=0.5, linetype=2) + 
#   geom_line(aes(date,est), alpha=0.6, lwd=0.75) + 
#   scale_colour_manual(values=c(col[5],col[4],col[3],col[1]))+
#   scale_fill_manual(values=c(col[5],col[4],col[3],col[1]))+ 
#   #ggtitle("Effect of choice of GAM smooth")+
#   #ylim(c(-17,21))+
#   theme_classic() + xlab("") + ylab("") 
# 
# bottom_panel <-ggarrange(panel.b, panel.c,widths=c(1.05,1), nrow=1,labels=c("B.", "C."))
# bottom_panel<-annotate_figure(bottom_panel, bottom = text_grob("Date"))
# 
# top_panel<-ggarrange(panel.a,labels=c("A."),nrow=1)
# 
# figure2<-ggarrange(top_panel,bottom_panel, nrow=2)
# figure2
# ggsave(figure2, filename = "Figure_S2_GAMtypes.png",bg="white", height = 8, width=8.5)
```

## Figures

::: {#suppfig-TSphos}
```{r}
#| echo: false
#| error: false
#| warning: false
print(tsphos)
```

Time series of total phosphorus in Lake Washington
:::

::: {#suppfig-TSbg}
```{r}
#| echo: false
#| error: false
#| warning: false
print(tsbg)
```

Time series of bluegreen algae (cyanobacteria) abundance in Lake Washington
:::

::: {#suppfig-CCF}
```{r}
#| echo: false
#| error: false
#| warning: false
ccf(dat$y_adj, dat$zp, main="CCF")
```

A CCF (Cross-Correlation Function) plot that represents the correlation between total phosphorus and blugreen algae abundance as a function of the lag (time difference) between them.
:::

::: {#suppfig-dlmSlope}
```{r}
#| echo: false
#| error: false
#| warning: false
DLM_plot
```

Time-varying slope of the relationship between bluegreen algae (response) and total phosphorus (driver) from a DLM
:::

::: {#suppfig-dlmFit}
```{r}
#| echo: false
#| error: false
#| warning: false
dlmFit
```

Fit of the DLM model with time-varying slope (black line) to observed bluegreen algae abundance (red points).
:::

::: {#suppfig-dlmresid}
```{r}
#| echo: false
#| error: false
#| warning: false
dlm_resid
```

Time series of residuals from the DLM model fit with a time-varying slope between blue-green algae and phosphorus
:::

::: {#suppfig-dlmqq}
```{r}
#| echo: false
#| error: false
#| warning: false
dlm_qq
```

Q-Q plot from the DLM model fit with a time-varying slope between blue-green algae and phosphorus
:::

::: {#suppfig-gamFit}
```{r, echo=FALSE}
pA
```

Fit of the GAM models (lines) to observed bluegreen algae (cyanobacteria) abundance (black points). Colors denote different specifications of the smoother functions.
:::

::: {#suppfig-gamresid}
```{r}
#| echo: false
#| error: false
#| warning: false
gam_resid
```

Time series of residuals from the five GAM models where residuals from models that do not have a temporally structured smooth term (s(Phos) and s(Phos, by = date)) have a trend through time.
:::

::: {#suppfig-gamqq}
```{r}
#| echo: false
#| error: false
#| warning: false
gam_qq
```

Q-Q for each of the five GAM models where residuals from models that do not have a temporally structured smooth term (s(Phos) and s(Phos, by = date)) have a trend through time.
:::

::: {#suppfig-gamSmooth}
```{r, echo=FALSE}
gam_smooth
```

Effect of the choice of smoother function on estimates of time-varying slope parameter where color denotes model type. 3 GAM smoother types (cr, tp, gp) are contrasted with a DLM (dlm).
:::

::: {#suppfig-gamknots}
```{r, echo=FALSE}
gamknots
```

Effect of smoothing degrees of freedom on the standard errors of the estimates of time-varying slope parameter where color denotes model type. We show 4 distinct knots (colors)
:::
