---
title: "NTL_LTER_TR Case Study"
format: html
editor: visual
---

## 

Approach for running LTER-NTL Trout Lake data using temporally structured GAMs and linear regression models with time periods defined a-priori.

```{r}
#| echo: false
#| include: false
#| error: false
#| warning: false

library(ggpubr)
library(nord)
library(ggplot2)
library(dplyr)
library(stringr)
library(dplyr)
library(nord)
library(tidyr)
library(data.table)
library(mgcv)
library(kableExtra)
library(knitr)
library(strucchange)
library(mgcv)

full_data <- readRDS("LTER_NTL_case_study/NTL_LTER_TL_data.rds")
analysis_data<-full_data%>%
  select(year4, mean_sec,mean_chl, Large,mean_totpuf,period)%>%
  mutate(mean_sec_scale=scale(mean_sec),mean_chl_scale=scale(mean_chl), 
         Large_scale=scale(Large),mean_totpuf_scale=scale(mean_totpuf))
```

First we start with looking at three time periods using linear regression models:
1. The historical regime with low water clarity 
2. A clear water regime where the introduction of Lake Trout into the system from stocking in 2006. 
3. A novel regime following the introduction of invasive, predatory water flea (*Bythotrephes*) in 2014 which lead to a reversion of water clarity to a less clear state.

To examine this we start by looking at a few key food web conditions using intercept only models through time:
1. Water clarity
2. Phosphorus - which impacts water clarity and is a common bottom-up process that could impact water clarity and we examine as an alternative hypothesis to the top down processes of Lake Trout and invasive speceis.
3. Abundance of large zooplankton *Daphnia* and *Calanoids*
4. Chlorophyll

```{r}
#| echo: false
#| warning: false
analysis_long<-analysis_data%>%
pivot_longer(!c(year4,period), names_to = 'Dataset', values_to='values')
dataset_names <- list(
  'mean_sec_scale'="Water Clarity",
  'mean_totpuf_scale'="Total Phosphorus",
  'mean_chl_scale'="Chlorophyll",
  'Large_scale'="Large Zooplankton"
)
facet_labeller <- function(variable,value){
  return(dataset_names[value])
}
ggplot(data=analysis_long%>%
         filter(Dataset=="mean_sec_scale"|Dataset=="mean_totpuf_scale"|
                 Dataset=="mean_chl_scale"| Dataset=="Large_scale"), aes(y=values, x=year4, group=Dataset, col=Dataset)) +
  facet_wrap(~Dataset,nrow=2,labeller=facet_labeller) +
  geom_hline(yintercept=0)+
  geom_point()+
  geom_line()+
  scale_colour_discrete(labels=facet_labeller)+
  geom_vline(xintercept=2007, lty=2)+
  geom_vline(xintercept=2014, lty=2)+
  xlab("Year")+
  guides(shape = FALSE)
```

We approach this by fitting a linear model with the a priori time periods as a factor and compare AIC to a single intercept model for wach variable.

```{r}
#| echo: false

WC_period<-lm(mean_sec_scale~as.factor(period), data=analysis_data)
WC_int<-lm(mean_sec_scale~1, data=analysis_data)

Large_period<-lm(Large_scale~as.factor(period), data=analysis_data)
Large_int<-lm(Large_scale~1, data=analysis_data)

chl_period<-lm(mean_chl_scale~as.factor(period), data=analysis_data)
chl_int<-lm(mean_chl_scale~1, data=analysis_data)

phos_period<-lm(mean_totpuf_scale~as.factor(period), data=analysis_data)
phos_int<-lm(mean_totpuf_scale~1, data=analysis_data)


AIC_table<-data.frame(Variable=c("Water Clarity", "Total Phosphorus", "Chlorophyll", "Large Zooplankton"), No.Period.AIC=round(c(AIC(WC_int),AIC(phos_int),AIC(chl_int),AIC(Large_int)),digits=2),
Period.AIC=round(c(AIC(WC_period),AIC(phos_period),AIC(chl_period),AIC(Large_period)),digits=2), Best.Model=c("Period", "No difference", "Period","Period"))


```
We find that water clarity, chlorophyll, and large zooplankton abundance are all better explained by a model that includes apriori defined time periods improves model fit. 

```{r}
#| echo: false

AIC_table %>%
  kbl() %>%
  kable_classic_2(full_width = F)
```

We have identified time-varying mean abundance or amount. We see that the relationship between large zooplankton and water clarity, and large zooplankton and chlorophyll, appears to be time-varying. Next we consider whether there are changing relationships between water clarity and these ecosystem dynamics by including a slope parameter and its interaction with time period. 

```{r}
#| echo: false
#| Include: false
#make a table 

zoop_slope_int<-lm(mean_sec~Large*as.factor(period), data=full_data)
chl_slope_int<- lm(mean_sec~mean_chl*as.factor(period), data=full_data)
phos_slope_int<- lm(mean_sec~scale(mean_totpuf)*as.factor(period), data=full_data)

zoop_slope<-lm(mean_sec~Large+as.factor(period), data=full_data)
chl_slope<- lm(mean_sec~mean_chl+as.factor(period), data=full_data)
phos_slope<- lm(mean_sec~scale(mean_totpuf)+as.factor(period), data=full_data)

Interaction<-data.frame(
  Covariates= c("Chlorophyll", "Total Phosphorus", "Large Zooplankton"),
  Period = round(c(AIC(chl_slope),AIC(phos_slope),AIC(zoop_slope)), digits=2),
  Interaction= round(c(AIC(chl_slope_int),AIC(phos_slope_int),AIC(zoop_slope_int)), digits=2)
)

Interaction %>%
  kbl() %>%
  kable_classic_2(full_width = F)

```
Both Chlorophyll and Large Zooplankton best explain water clarity with a time varying relationship. We examine how these relationships change through time. 

```{r}
#| echo: false
print("Chlorophyll")
CHLresults<-data.table(coef(summary(chl_slope_int)), keep.rownames = 'term')
CHLresults %>%
  kbl() %>%
  kable_classic_2(full_width = F)

print("Zooplankton")
zoopresults<-data.table(coef(summary(zoop_slope_int)), keep.rownames = 'term')
zoopresults %>%
  kbl() %>%
  kable_classic_2(full_width = F)

Fit_zoop<-ggplot(data=analysis_long%>%
         filter(Dataset=="mean_sec"|
                  Dataset=="Large_scale")%>%
  pivot_wider(names_from = Dataset, values_from = values), aes(group=period,col=period, y=mean_sec,x=Large_scale)) +
  stat_smooth(method = "lm")+
  scale_colour_discrete(labels=c("1980 - 2006", "2007 - 2014", "2014 - 2022"))+
  geom_point()+
  ylab("Water Clarity")+
  xlab("Large Zooplankton")

TS_zoop<- ggplot(data=analysis_long%>%
         filter(Dataset=="mean_sec_scale"|
                  Dataset=="Large_scale"), aes(y=values, x=year4, group=Dataset, col=Dataset)) +
  #facet_wrap(~area1,scales="free_y") +
  geom_point()+
  geom_line()+
  scale_colour_discrete(labels=c("Water Clarity", "Large Zooplankton"))+
  geom_vline(xintercept=2007, lty=2)+
  geom_vline(xintercept=2014, lty=2)+
  guides(shape = FALSE)+
  xlab("Year")



Fit_chl <-ggplot(data=analysis_long%>%
         filter(Dataset=="mean_sec"|
                  Dataset=="mean_chl_scale")%>%
  pivot_wider(names_from = Dataset, values_from = values), aes(group=period,col=period, y=mean_sec,x=mean_chl_scale)) +
  stat_smooth(method = "lm")+
  scale_colour_discrete(labels=c("1980 - 2006", "2007 - 2014", "2014 - 2022"))+
  geom_point()+
  ylab("Water Clarity")+
  xlab("Chlorophyll")

TS_chl<-ggplot(data=analysis_long%>%
         filter(Dataset=="mean_sec_scale"|
                  Dataset=="mean_chl_scale"), aes(y=values, x=year4, group=Dataset, col=Dataset)) +
  #facet_wrap(~area1,scales="free_y") +
  geom_point()+
  geom_line()+
  scale_colour_discrete(labels=c("Water Clarity", "Chlorophyll"))+
  geom_vline(xintercept=2007, lty=2)+
  geom_vline(xintercept=2014, lty=2)+
  guides(shape = FALSE)+
  xlab("Year")

TS_zoop
Fit_zoop
TS_chl
Fit_chl



```

If we take the Zooplankton-water quality relationship and fit a temporally structured gam and compare it to the linear model, we find similar results. Both predictions also acccurately identify 2006 as a breakpoint and a second break point in 2016.

```{r}
#| echo: false

prediction=data.frame(prediction=predict(zoop_slope_int),year=unique(analysis_long$year4))
ggplot(data=prediction,aes(y=prediction, x=year)) +
  #facet_wrap(~area1,scales="free_y") +
 # geom_point()+
  geom_point(data=analysis_long%>%
         filter(Dataset=="mean_sec"), aes(y=values, x=year4, group=Dataset, col=Dataset))+
  geom_line()+
  ggtitle('Linear Model')+
  geom_vline(xintercept=2007, lty=2)+
  geom_vline(xintercept=2014, lty=2)+
  guides(shape = FALSE)

y.ts <- ts(data=prediction, frequency=1)
# fit breakpoint model
bp.y <- breakpoints(y.ts ~ 1)
summary(bp.y) 
prediction[26,]
prediction[36,]
gam1<-gam(mean_sec~s(year4, by=Large,k=10), data=full_data)
#plot(gam1)
gam2<-gam(mean_sec~s(year4,k=10), data=full_data)
gam3<-gam(mean_sec~s(Large,k=10), data=full_data)

summary(gam2)
prediction2=data.frame(prediction=predict(gam1),year=unique(analysis_long$year4))
ggplot(data=prediction2,aes(y=prediction, x=year)) +
  #facet_wrap(~area1,scales="free_y") +
  #geom_point()+
  geom_line()+
    geom_point(data=analysis_long%>%
         filter(Dataset=="mean_sec"), aes(y=values, x=year4, group=Dataset, col=Dataset))+
  ggtitle('GAM')+
  geom_vline(xintercept=2007, lty=2)+
  geom_vline(xintercept=2014, lty=2)+
  guides(shape = FALSE)

y.ts <- ts(data=prediction2, frequency=1)
# fit breakpoint model
bp.y <- breakpoints(y.ts ~ 1)
summary(bp.y) 
prediction2[26,]
prediction2[36,]

```

Looking at a 10 year rolling window where the first and last year of the time series changes and see what the slop estimate is as a function of those? Maybe also do a 5-year and a 15-year to also show a function of time series length?

```{r}
#| echo: false

results <- data.frame()
important_variables<-data.frame()
sf_fun <- function(first_year, window){
for(i in 1:length(window)){
 for(j in 1:length(first_year)){
 the_data <- analysis_long%>%
         filter(Dataset=="mean_sec"|
                  Dataset=="mean_chl_scale")%>%
  pivot_wider(names_from = Dataset, values_from = values)%>% 
   #filter(year4 <= first_year[j]+window[i] & year4 >= first_year[j])
   filter(year4 >= first_year[j]-window[i] & year4 <= first_year[j])
 
  the_lm <-  lm(mean_sec~mean_chl_scale, data=the_data)
  
 
 # slope_tibble[["last_year"]] <- last_year
 # slope_tibble[["first_year"]] <- first_year
  important_variables_temp <- cbind(Intercept=summary(the_lm)$coefficients[1,1],
                              Slope=summary(the_lm)$coefficients[2,1], 
                              last_year=first_year[j], 
                              first_year=first_year[j]-window[i],
                              window=window[i])
  
important_variables <-rbind(important_variables_temp, important_variables)
 #rbind(slope_tibble, important_variables)  
 } 
results <- rbind(important_variables, results)
}
return(results)
}
window<-c(3,5,7,9,11,13,15)
results_window<-sf_fun(2023:1992,window)
p <- ggplot(results_window, aes(x=last_year, y=as.factor(window), fill= Slope)) + 
  geom_tile()+
  xlab("Last Year")+
  ylab("Window")+
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000", midpoint = 0)
p 

p <- ggplot(results_window, aes(x=last_year, y=as.factor(window), fill= Intercept)) + 
  geom_tile()+
  xlab("Last Year")+
  ylab("Window")+
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000", midpoint = 6)
#p
```



Looking at GAMs for Chlorophyll

```{r}
#| echo: false
#| output: false


results <- data.frame()
important_variables<-data.frame()
sf_fun_gam <- function(first_year, window){
for(i in 1:length(window)){
 for(j in 1:length(first_year)){
 the_data <- analysis_long%>%
         filter(Dataset=="mean_sec"|
                  Dataset=="mean_chl_scale")%>%
  pivot_wider(names_from = Dataset, values_from = values)%>% 
   #filter(year4 <= first_year[j]+window[i] & year4 >= first_year[j])
   filter(year4 >= first_year[j]-window[i] & year4 <= first_year[j])
 
  the_gam <-  gam( mean_sec~s(year4, by=mean_chl_scale,k=6), data=the_data)
  
plot_cr <- plot(the_gam, seWithMean = TRUE, n = nrow(the_data))

df_cr <- data.frame(date = the_data$year4, 
                    est = plot_cr[[1]]$fit, 
                    se = plot_cr[[1]]$se, type="cr")
 # slope_tibble[["last_year"]] <- last_year
 # slope_tibble[["first_year"]] <- first_year
  important_variables_temp <- cbind(est=mean(df_cr$est), 
                              last_year=first_year[j], 
                              first_year=first_year[j]-window[i],
                              window=window[i])
  
important_variables <-rbind(important_variables_temp, important_variables)
 #rbind(slope_tibble, important_variables)  
 } 
results <- rbind(important_variables, results)
}
return(results)
}
window<-c(7,9,11,13,15)
results_window<-sf_fun_gam(2023:1992,window)


```

```{r}
#| echo: false

p <- ggplot(results_window, aes(x=last_year, y=as.factor(window), fill= est)) + 
  geom_tile()+
  xlab("Last Year")+
  ylab("Window")+
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000", midpoint = 0)
p 

```

Same process with Zooplankton

```{r}
#| echo: false

results <- data.frame()
important_variables<-data.frame()
sf_fun_zoop <- function(first_year, window){
for(i in 1:length(window)){
 for(j in 1:length(first_year)){
 the_data <- analysis_long%>%
         filter(Dataset=="mean_sec"|
                  Dataset=="Large_scale")%>%
  pivot_wider(names_from = Dataset, values_from = values)%>% 
   #filter(year4 <= first_year[j]+window[i] & year4 >= first_year[j])
   filter(year4 >= first_year[j]-window[i] & year4 <= first_year[j])
 
  the_lm <-  lm(mean_sec~Large_scale, data=the_data)
  
 
 # slope_tibble[["last_year"]] <- last_year
 # slope_tibble[["first_year"]] <- first_year
  important_variables_temp <- cbind(Intercept=summary(the_lm)$coefficients[1,1],
                              Slope=summary(the_lm)$coefficients[2,1], 
                              last_year=first_year[j], 
                              first_year=first_year[j]-window[i],
                              window=window[i])
  
important_variables <-rbind(important_variables_temp, important_variables)
 #rbind(slope_tibble, important_variables)  
 } 
results <- rbind(important_variables, results)
}
return(results)
}
window<-c(3,5,7,9,11,13,15)
results_window<-sf_fun_zoop(2023:1992,window)
p <- ggplot(results_window, aes(x=last_year, y=as.factor(window), fill= Slope)) + 
  geom_tile()+
  xlab("First Year")+
  ylab("Window")+
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000", midpoint = 0)
p 

p <- ggplot(results_window, aes(x=last_year, y=as.factor(window), fill= Intercept)) + 
  geom_tile()+
  xlab("Last Year")+
  ylab("Window")+
  scale_fill_gradient2(low = "#075AFF",
                       mid = "#FFFFCC",
                       high = "#FF0000", midpoint = 6)
#p
```