---
title: "Appendix 3: Gulf of Alaska salmon analysis"
format: pdf
editor: source
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S3.
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
    - kind: float
      key: supptab
      latex-env: supptab
      reference-prefix: Table S3.
      space-before-numbering: false
      latex-list-of-description: Supplementary Table
---

# Appendix 3: Gulf of Alaska salmon analysis

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  warning = FALSE,  
  error = FALSE,    
  message = FALSE
)
library(paletteer)
library(MARSS)
library(mgcv)
library(dplyr)
library(forecast)
library(ggplot2)
library(lubridate)
library(marssTMB)
library(tidyverse)
library(strucchange)
library(data.table)
library(DHARMa)
library(modelr)
library(forecastHybrid)
library(ggpubr)
```

Here we show the LM, GAM, and DLM analysis for the Gulf of Alaska salmon dataset. For complete code, see the [full quarto document](https://github.com/ecosystem-state/nonstationarity-best-practices/blob/main/GOA_salmon_case_study/Appendix3_GoASalmon.qmd)

## Data Processing

First, we standardize and examine the trends in SST and salmon abundace from Litzow et al. 2018 updated with data through 2025 (@suppfig-TSsalmon & @suppfig-TSsst).

```{r, echo=FALSE}
# Load salmon timeseries collapsed into DFA trend
salm.dat <- readRDS(here::here("03_GOA_salmon_case_study/Data/dfa_trend.rds")) 

# Load 3-year running mean winter SST timeseries
sst.dat <- readRDS(here::here("03_GOA_salmon_case_study/Data/winterSST_3yr_running_mean.rds")) 

# Bind
right_join(salm.dat, sst.dat) -> dat

# Plot salmon DFA
salmon_DFA<- ggplot(dat, aes(year, salmon_DFA)) +
  theme_classic() +
  geom_line() +
  geom_hline(yintercept = 0) +
  geom_point() +
  geom_errorbar(aes(x=year, ymin=conf.low, ymax=conf.high)) +
  xlab("") +
  ylab("Salmon Trend")

# Plot sst
SST_TS<- ggplot(dat, aes(year, sst_3yr_running_mean)) +
  theme_classic() +
  geom_line() +
  geom_point() +
  xlab("") +
  ylab("3-year running mean SST")

# Scale predictor sst
dat %>%
  mutate(z_sst = scale(sst_3yr_running_mean)) %>%
  na.omit() -> dat2
```

## LM

We start our investigation using linear models, to evaluate the relationship between salmon catch and SST anomalies with periods defined *a-priori* as before and after 1988/1989 (from Litzow et al. 2018). Previous investigations used this approach, and while we primarily discuss DLMs and GAMs for this dataset in the main text, we include this illustrative example here for completeness and comparability to the previous study.

We compare linear model structures using classic model selection criteria, AIC. First, we examine a 1) intercept only model, 2) salmon abundance with period as a predictor, 3) salmon abundance with period and SST anomalies as predictors, and 4) an interaction between SST anomalies and period. This shows that model 2 is a better fit to the data than model 1 indicating that a time-dependent intercept is an improvement to model fit than a time-independent intercept. Similarly, model 4 is the best model showing that a time-dependent slope with SST is an improvement to a time-independent slope (@supptab-AIC).

```{r, echo=FALSE}
dat2 %>%
  mutate(period = case_when((year <= 1988) ~ "early",
                            (year > 1988) ~ "late")) -> dat2

mod_period<-lm(salmon_DFA~as.factor(period), data=dat2) # with period
mod_int<-lm(salmon_DFA~1, data=dat2) # intercept only
mod_sst.period.int<-lm(salmon_DFA~z_sst*as.factor(period), data=dat2) # with interaction
mod_sst.period<-lm(salmon_DFA~z_sst+as.factor(period), data=dat2) # no interaction
mod_sst.period.noint<-lm(salmon_DFA~0+z_sst:as.factor(period), data=dat2) # no interaction


mod_int_AIC <- AIC(mod_int)
mod_mod_period <- AIC(mod_period) 
mod_sst.period_AIC <- AIC(mod_sst.period)
mod_sst.period.int_AIC <- AIC(mod_sst.period.int) 
mod_sst.period.noint_AIC <- AIC(mod_sst.period.noint) 

model_names <- c("1. Intercept only", "2. Intercept by period only \n (time-dependent intercept)", 
"3. Intercept by period and SST", "4. Interaction between SST and period \n (time-dependent slope and intercept)", "5. Time-dependent slope")
AIC <- c(mod_int_AIC, mod_mod_period, mod_sst.period_AIC,mod_sst.period.int_AIC,mod_sst.period.noint_AIC)
AIC_tab <-data.frame(Model = model_names, AIC = round(AIC, 2))

```

Functionally, model 4 has a time-dependent slope and a time-dependent intercept which we can see by both plotting the relationships and examining the model output. We can see a strong positive correlation prior to 1988 (@suppfig-TSslm) followed by a weak negative relationship which is represented in the slopes of the relationship (@supptab-results & @suppfig-fitslm).

```{r, echo = FALSE}

mod.results <-data.table(round(coef(summary(mod_sst.period.int)),2), keep.rownames = 'term')

```

```{r, echo = FALSE}


Fit_slm<-ggplot(data=dat2, aes(group=period,col=period, y=salmon_DFA,x=z_sst)) +
  stat_smooth(method = "lm")+
  scale_colour_discrete(labels=c("before 1988/89", "after 1988/89"))+
  geom_point()+
  ylab("Salmon trend")+
  xlab("SST")+ theme_classic()

TS_slm<- ggplot()+
  geom_line(dat2, mapping = aes(x=year, y=scale(salmon_DFA)), color = "salmon", size = 1) +
  geom_line(dat2, mapping = aes(x = year, y = z_sst), color = "darkblue", size = 1)+
  geom_point(dat2, mapping = aes(x=year, y=scale(salmon_DFA)), color = "salmon") +
  geom_point(dat2, mapping = aes(x = year, y = z_sst), color = "darkblue")+
  scale_colour_discrete(labels=c("salmon trend", "SST"))+
  geom_vline(xintercept=1988, lty=2)+
  xlab("Year")+ theme_classic()

```

## GAM

Next, we consider generalized additive models. We fit two distinct GAM models to the salmon catch data and SST anomalies to illustrate the differences in a model with a time-dependent slope and a time-dependent intercept. We find that when fitting a time-dependent slope and time-dependent intercept model, most of the variability in the relationship is explained by the time-dependent intercept (@suppfig-tvintgam) while the time-dependent slope does not change much through time (@suppfig-tvslopegam). When only a time-dependent slope is fit to the data, the slope parameter is much more variable through time (@suppfig-tvslopegam), and looks similar to the slope estimated by the linear models (@suppfig-fitslm).

Next, we check the residuals plots for both parameterizations of the models. The time-dependent slope model appears to have some residual temporally variability while there are no trends through time in the model with both a time-dependent slope and a time-dependent intercept (@suppfig-gamresid) and the qq-pot looks reasonable for both (@suppfig-gamqq).

```{r,include=FALSE}
dat_aksalmon <- readRDS("../03_GOA_salmon_case_study/data_for_AK_salmon_example.rds")
col<-paletteer_d("nationalparkcolors::Everglades")
#renaming
dat_aksalmon <- dplyr::rename(dat_aksalmon,
                              response = salmon_DFA,
                              driver = z_sst, 
                              date = year)
K_AK <- 9#round(0.99*nrow(dat_aksalmon))
dat_aksalmon$n_date <- seq(1,nrow(dat_aksalmon))

gam_aksalmon_1 <- gam(response ~ s(n_date,k=K_AK) + s(n_date, by = driver, bs = "gp",k=K_AK), data = dat_aksalmon)
gam_aksalmon_2 <- gam(response ~ s(n_date, by = driver, bs = "gp",k=K_AK), data = dat_aksalmon)
gam_aksalmon_3 <-  gam(response ~ s(driver, bs = "gp",k=K_AK), data = dat_aksalmon)

# extract the smooth estimates with SEs to compare to DLM
plot_3 <- plot(gam_aksalmon_1, seWithMean = TRUE, n = nrow(dat_aksalmon))
plot_4 <- plot(gam_aksalmon_2, seWithMean = TRUE, n = nrow(dat_aksalmon))
summary(gam_aksalmon_2)
coef_gam_AK1 <- data.frame(date = dat_aksalmon$date,
                        int_est = plot_3[[1]]$fit,
                        int_se = plot_3[[1]]$fit,
                        slope_est = plot_3[[2]]$fit,
                        slope_se = plot_3[[2]]$fit,
                        time_varying = "Intercept + slope",
                        dataset = "Alaska salmon : GAM")
coef_gam_AK2 <- data.frame(date = dat_aksalmon$date,
                        int_est = NA,
                        int_se = NA,
                        slope_est = plot_4[[1]]$fit,
                        slope_se = plot_4[[1]]$se,
                        time_varying = "Slope",
                        dataset = "Alaska salmon : GAM")

coefs_gam <- rbind(coef_gam_AK1, coef_gam_AK2)

slope_panel_gam <-ggplot(coefs_gam, aes(date, slope_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=slope_est-2*slope_se, ymax = slope_est+2*slope_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-dependent slope") + 
  xlab("") + theme_classic() + 
  ggtitle("Slope")+
  #ylim(c(-5,11))+
  facet_wrap(~ dataset, scale="free") + 
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  #guides(fill="none",colour="none",lty="none")+
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )
coefs_gam$int_est[is.na(coefs_gam$int_est)] <- 1.2863
coefs_gam$int_se[is.na(coefs_gam$int_est)] <- 0.2398

int_panel_gam <- ggplot(coefs_gam, aes(date, int_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=int_est-2*int_se, ymax = int_est+2*int_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-dependent intercept") + 
  xlab("") + theme_classic() + 
  facet_wrap(~ dataset, scale="free") + 
  ggtitle("Intercept")+
  #ylim(-20,10.5)+
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )


```

```{r echo = FALSE}
resid_df1 <- data.frame(resid = resid(gam_aksalmon_1),
                       date = dat_aksalmon$date,
                       Model = "Intercept + slope")
resid_df2 <- data.frame(resid = resid(gam_aksalmon_2),
                       date = dat_aksalmon$date,
                       Model = "Slope")
resid_df3 <- data.frame(resid = resid(gam_aksalmon_3),
                       date = dat_aksalmon$date,
                       Model = "Intercept")
p_values<-data.frame(p_value=c(
                     summary(gam(resid(gam_aksalmon_1)~s(dat_aksalmon$date)))$s.pv,
                     summary(gam(resid(gam_aksalmon_2)~s(dat_aksalmon$date)))$s.pv,
                     summary(gam(resid(gam_aksalmon_3)~s(dat_aksalmon$date)))$s.pv), 
           Model=c("Intercept + slope",  "Slope", "Intercept"))
resids <- rbind(resid_df1, resid_df2)%>%#,resid_df3)
left_join(p_values)

gamresid <- ggplot(resids, aes(date, resid)) + 
  geom_point() + 
  geom_smooth(method = 'gam',formula = y ~ s(x)) + 
  theme_classic() + xlab("Date") + ylab("Residuals") +
     geom_text(
    aes(x = 1970, y = -2, label =paste0("p = ",round(p_value,4))),
    inherit.aes = FALSE)+
  ylim(c(-3, 4))+
  facet_wrap(~ Model, nrow=2) + ggtitle("GAM Residuals")
```

```{r echo = FALSE}

simulationOutputcr1 <- simulateResiduals(fittedModel =gam_aksalmon_1)
simulationOutputcr2 <- simulateResiduals(fittedModel =gam_aksalmon_2)

# gamquant<-ggplot(resids, aes(sample = resid)) + stat_qq() + stat_qq_line() + labs(
#     x = "Theoretical Quantiles",
#     y = "Sample Quantiles"
#   ) + theme_classic() + facet_wrap(~ Model,scale="free_y") + ggtitle("GAM Residuals")

#simulationOutput <- simulateResiduals(fittedModel = gam_aksalmon_3, refit = T)
#plot(simulationOutput)
#testDispersion(simulationOutput)
#plotResiduals(simulationOutput, dat_aksalmon$n_date)
#plotResiduals(simulationOutput, dat_aksalmon$driver)
```

```{r echo = FALSE}
#k-fold cross validation
cv_dat<-dat_aksalmon%>%
  mutate(id = row_number())

train_df = slice_sample(cv_dat,n=25) #size = 0.5)
test_df = anti_join(cv_dat, train_df, by = "id")

folds_plot<-ggplot(train_df, aes(x = date, y = response)) + 
  geom_point(aes(color="Train Data")) + 
  geom_point(data = test_df, aes(color="Test Data"))+
    scale_color_manual(values=c( "red","black"))+
  theme_classic()

Timevarying_Intercept_and_Slope <- gam(response ~ s(n_date,k=K_AK) + s(n_date, by = driver, bs = "gp",k=K_AK), data = train_df)
Timevarying_Slope <- gam(response ~ s(n_date, by = driver, bs = "gp",k=K_AK), data = train_df)
Not_Timevarying <-  gam(response ~ s(driver, bs = "gp",k=K_AK), data = train_df)


foldsbymodelplot<-train_df |> 
  gather_predictions(Timevarying_Intercept_and_Slope,
                     Timevarying_Slope) |>
                     #Not_Timevarying) |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = date, y = response)) + 
  geom_point() + 
  geom_line(aes(y = pred), color="red") + 
     # scale_color_manual(values=c("black","red"))+
  #geom_point(data = test_df, color = "red")+
  facet_wrap(~model)+
  theme_classic()
```

We need to decide if there is enough ecological evidence to support a model with both a time-dependent slope and time-dependent intercept and whether that model has signs of overfitting. We can do this by using cross validation and comparing the root mean square error (RMSE) of a set of training data compared to test data. For time series data its important that cross validation methods retain the temporal structure of the data. To do this, we used approximate leave-future out cross validation (LFO-CV) where we fit the model to 15 years of data (training data, first year to year n-1) and predict 1 year ahead (test data, year n). The model is used to produce in-sample predictions for the training data and one-year-ahead out of sample predictions for the test data. Then the model is iteratively refit walking one year ahead such that the training data then includes n and the test data is n+1. RMSE is then calculated for both the training and the test datasets based on predictions for each iteration. If RMSE of the training data is significantly lower than RMSE of the test data, it indicates the model is likely overfitting. 

We compared LFO-CV for the training and test data for both candidate models: time-dependent slope and intercept, and time-dependent slope only (@suppfig-lfofold). For the time-dependent slope and intercept model, the RMSE of the training data was significantly lower (p < 0.05) than the RMSE of the test data (@supptab-rmse). The RMSE of the training data was lower than the RMSE for the test data for the time-dependent slope model, but they were not significantly different (p > 0.05). Collectively with the residual diagnostics, we conclude that the time-dependent slope model is appropriate for interpreting ecological dynamics and performs better than the model that also included a time-dependent intercept (based on evidence of overfitting with this model parameterization). However, if the goal of fitting this model was prediction, it would be valuable to address the residual temporal variability that is not explained by SST and the time-dependent slope. Additional predictors and considering additional extrinsic factors that may contribute to the time-dependent salmon catch - SST relationship would be the next step.

```{r echo = FALSE, include=FALSE}
train_df_pred_TIS<-train_df |> 
  gather_predictions(Timevarying_Intercept_and_Slope) 
  train_kfold_tis<-sqrt(sum((train_df_pred_TIS$pred-train_df_pred_TIS$response)^2)/nrow(train_df_pred_TIS))

train_df_pred_TS<-train_df |> 
  gather_predictions(Timevarying_Slope) 
  train_kfold_ts<-sqrt(sum((train_df_pred_TS$pred-train_df_pred_TS$response)^2)/nrow(train_df_pred_TS))
    
test_df_pred_TIS<-test_df |> 
  gather_predictions(Timevarying_Intercept_and_Slope) 
  test_kfold_tis<-sqrt(sum((test_df_pred_TIS$pred-test_df_pred_TIS$response)^2)/nrow(test_df_pred_TIS))

test_df_pred_TS<-test_df |> 
  gather_predictions(Timevarying_Slope) 
  test_kfold_ts<-sqrt(sum((test_df_pred_TS$pred-test_df_pred_TS$response)^2)/nrow(test_df_pred_TS))
    
```

```{r echo = FALSE, include=FALSE}
dat<-dat_aksalmon
pred<- data.frame()
min_years <- 9
full_length <-nrow(dat)
jstart<- 1
K_AK<-8

for (j in (jstart-1):(full_length-min_years-1)) {
    train_index <- (jstart):(min_years+j)# All indices except the j-th
    test_index <- min_years+j+1                # The j-th index
    train_data<-dat[which(dat$date %in% unique(dat$date)[train_index]), ]
    
    # Fitting the 3 models
    gam_model_1 <- gam(response  ~ s(n_date,k=K_AK) + s(n_date, by = driver, bs = "gp",k=K_AK),data=train_data )
    gam_model_2 <- gam(response ~ s(n_date, by = driver, bs = "gp",k=K_AK), data  = train_data)
    gam_model_3 <-gam(response ~ s(driver, bs = "gp",k=K_AK), data  = train_data)
    
    #RMSE for training data for each model
    rmse_train1<- sqrt(sum((predict(gam_model_1)-train_data$response)^2)/nrow(train_data))
    rmse_train2<- sqrt(sum((predict(gam_model_2)-train_data$response)^2)/nrow(train_data))
    rmse_train3<- sqrt(sum((predict(gam_model_3)-train_data$response)^2)/nrow(train_data))

    
    newdata<-dat[which(dat$date %in% unique(dat$date)[test_index]), ]
    gam1_pred=predict(gam_model_1, newdata = newdata)
    gam2_pred=predict(gam_model_2, newdata = newdata)
    gam3_pred=predict(gam_model_3, newdata = newdata)
    rmse_test1=sqrt(sum((gam1_pred-newdata$response)^2)/nrow(newdata))
    rmse_test2=sqrt(sum((gam2_pred-newdata$response)^2)/nrow(newdata))
    rmse_test3=sqrt(sum((gam3_pred-newdata$response)^2)/nrow(newdata))
    results=cbind(rmse_train1, rmse_test1,gam1_pred,
                             rmse_train2, rmse_test2,gam2_pred,
                             rmse_train3, rmse_test3,gam3_pred,
                             date=newdata$date)
    # Predict the excluded observation
    pred <- rbind(pred,results)
  }

train_dat_pred<-left_join(dat,pred)
rmse_pred<-apply(pred, 2, mean)
t.test(pred$rmse_train1,pred$rmse_test1)     
t.test(pred$rmse_train2,pred$rmse_test2)     
t.test(pred$rmse_train3,pred$rmse_test3)    
```

```{r echo = FALSE}
tvintsl_lfo<-  ggplot(data=train_dat_pred,aes(x = date, y = response)) + 
  geom_point() + 
    ylim(c(-6,8))+
  ggtitle("Time-dependent slope and intercept")+
  geom_line(aes(y = gam1_pred), color = "red") + 
 # facet_wrap(~model)+
  theme_classic()
  
tvsl_lfo<-  ggplot(data=train_dat_pred,aes(x = date, y = response)) + 
  geom_point() + 
    ylim(c(-6,8))+
    ggtitle("Time-dependent slope")+
  geom_line(aes(y = gam2_pred), color = "red") + 
 # facet_wrap(~model)+
  theme_classic()
    
sl_lfo<-  ggplot(data=train_dat_pred,aes(x = date, y = response)) + 
  geom_point() + 
  ylim(c(-6,8))+
      ggtitle("Time-independent slope")+
  geom_line(aes(y = gam3_pred), color = "red") + 
 # facet_wrap(~model)+
  theme_classic()

```


```{r echo = FALSE}
#RMSE tables
names<- c("leave-future-out train", "leave-future-out test")
models <- c("Time-dependent slope & intercept", "Time-dependent slope")
row1<- c(rmse_pred[1],rmse_pred[2])
row2<- c(rmse_pred[4],rmse_pred[5])
cv<-data.frame(rbind(round(row1,2),round(row2,2)))
rownames(cv)=models
colnames(cv)=names
```
\break

## DLM

```{r,include=FALSE}


dat <- dat_aksalmon
m <- 2
TT <- nrow(dat)
B <- diag(m)  ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)  ## 2x2; all 0 for now
diag(Q)[1] <- c("q.alpha")
diag(Q)[2] <- c("q.beta")
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1, 1, ] <- rep(1, TT)  ## Nx1; 1's for intercept
Z[1, 2, ] <- dat$driver  ## Nx1; predictor variable
A <- matrix(0)  ## 1x1; scalar = 0
R <- matrix("r")  ## 1x1; scalar = r
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))
## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
# convert response to matrix
dat_mat <- matrix(dat$response, nrow = 1)
# fit the model -- crank up the maxit to ensure convergence
dlm_aksalmon_1 <- MARSS(dat_mat, inits = inits_list, model = mod_list,
               control = list(maxit=4000))

# Fit a second model with just time-dependent slope
diag(Q)[1] <- 1e-10
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
dlm_aksalmon_2 <-MARSS(dat_mat, inits = inits_list, model = mod_list,
               control = list(maxit=4000))

coef_dlm_AK1 <- data.frame(date = dat_aksalmon$date,
                        int_est = dlm_aksalmon_1$states[1,],
                        int_se = dlm_aksalmon_1$states.se[1,],
                        slope_est = dlm_aksalmon_1$states[2,],
                        slope_se = dlm_aksalmon_1$states.se[2,],
                        time_varying = "Intercept + slope",
                        dataset = "Alaska salmon : DLM")
coef_dlm_AK2 <- data.frame(date = dat_aksalmon$date,
                        int_est = NA,
                        int_se = NA,
                        slope_est = dlm_aksalmon_2$states[2,],
                        slope_se = dlm_aksalmon_2$states.se[2,],
                        time_varying = "Slope",
                        dataset = "Alaska salmon : DLM")

coefs <- rbind(coef_dlm_AK1, coef_dlm_AK2)

slope_panel <-ggplot(coefs, aes(date, slope_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=slope_est-2*slope_se, ymax = slope_est+2*slope_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-dependent slope") + 
  xlab("") + theme_classic() + 
  ggtitle("Slope")+
  #ylim(c(-5,11))+
  facet_wrap(~ dataset, scale="free") + 
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
 # guides(fill="none",colour="none",lty="none")+
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )

int_panel <- ggplot(coefs, aes(date, int_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=int_est-2*int_se, ymax = int_est+2*int_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-dependent intercept") + 
  xlab("") + theme_classic() + 
  facet_wrap(~ dataset, scale="free") + 
  ggtitle("Intercept")+
  #ylim(-20,10.5)+
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-dependent") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )


```

We fit two distinct DLM models to the salmon catch data and SST anomalies to compare to the results of the GAM models. Generally, the parameters estimated by the DLM are more certain and follow similar patterns through time as the GAM (@suppfig-tvint & @suppfig-tvslope). There is also more interannual variability in the parameter estimates of the DLM compared to the continuously smoothed pattern of the GAM. We also find similar residual patterns for the DLM compared to the GAM (@suppfig-dlmresid & @suppfig-dlmqq).

```{r echo=FALSE, include=FALSE, warning=FALSE}
pred <- predict(dlm_aksalmon_1)$pred
pred2 <- predict(dlm_aksalmon_2)$pred

resid_df1 <- data.frame(resid =residuals(dlm_aksalmon_1)$.std.resids, #pred$y - pred$estimate,
                       date = dat_aksalmon$date,
                       Model = "Intercept + slope")
resid_df2 <- data.frame(resid = residuals(dlm_aksalmon_2)$.std.resids, #pred2$y - pred2$estimate,
                       date = dat_aksalmon$date,
                       Model = "Slope")

p_values<-data.frame(p_value=c(summary(gam(residuals(dlm_aksalmon_1)$.std.resids~s(dat_aksalmon$date)))$s.pv,
                     summary(gam(residuals(dlm_aksalmon_2)$.std.resids~s(dat_aksalmon$date)))$s.pv), 
           Model=c("Intercept + slope","Slope"))

resids <- rbind(resid_df1, resid_df2)%>%
  left_join(p_values)

dlm_resids<-ggplot(resids, aes(date, resid)) + 
  geom_point() + 
  ylim(c(-2,4))+
  geom_smooth(method = 'gam',formula = y ~ s(x)) +
     geom_text(
    aes(x = 1980, y = 3.5, label =paste0("p = ",round(p_value,4))),
    inherit.aes = FALSE)+
  theme_classic() + xlab("Date") + ylab("Residuals") + 
  facet_wrap(~ Model, nrow=2, scale="free_y") + ggtitle("DLM Residuals")
```

```{r}
#| echo: false
#| include: false
#| error: false
#| warning: false
resid<-residuals(dlm_aksalmon_1)
resid_dlm <- predict(dlm_aksalmon_1)$pred
resid_dlm$resid <- resid_dlm$y - resid_dlm$estimate
dat$n_date <- as.numeric(as.factor(dat_aksalmon$date))
resid_dlm$date <- dat_aksalmon$date

p<-summary(gam(resid_dlm$resid~s(dat_aksalmon$date)))$s.pv
dlm_resid1<- ggplot(resid_dlm, aes(date, resid)) + 
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x)) + 
  theme_classic() + xlab("Date") + ylab("Residuals") + ggtitle("DLM Residuals")+
  annotate("text", x = 1980, y = -3, label =paste0("Temporal Trend: p = ",round(p,3)),color = "black", size = 5)+
  annotate("text", x = 1980, y = -3, label =paste0("Temporal Trend: p = ",round(p,3)),color = "black", size = 5)

## KS Test
ks<- ks.test(
  x = resid$.std.resids,
  y = "pnorm", 
  mean = mean(resid$.std.resids, na.rm = TRUE), 
  sd = sd(resid$.std.resids, na.rm = TRUE)
)$p.value

# Dispersion test
clean_resid <- na.omit(resid$.std.resids)
n_obs <- length(clean_resid)                  # Number of data points
param_count <- dlm_aksalmon_1$num.params                # Total estimated parameters
df_residual <- n_obs - param_count            # Degrees of freedom
chi_stat <- sum(clean_resid^2)
p_value <- pchisq(chi_stat, df = df_residual, lower.tail = FALSE)
cat("Chi-Square Stat:", chi_stat, "\n",
    "Degrees of Freedom:", df_residual, "\n",
    "P-value for Overdispersion:", p_value, "\n")

# outlier
res_obj <- MARSSresiduals(dlm_aksalmon_1)
obs_res <- res_obj$std.residuals[1, ] # Extract for the first response variable
obs_res <- obs_res[!is.na(obs_res)]   # Remove missing values
n_time <- length(obs_res)
alpha_corrected <- 0.05 / n_time
z_critical <- qnorm(1 - alpha_corrected / 2)
total_outliers <- sum(abs(obs_res) > z_critical)
prob_outlier <- alpha_corrected
binomial_test <- binom.test(
  x = total_outliers,
  n = n_time,
  p = prob_outlier,
  alternative = "greater" # Testing if we have MORE outliers than expected
)

dlm_qq_plot1 <-ggplot(resid, aes(sample = .std.resids)) + 
  stat_qq() + stat_qq_line(col='red') + labs(
     x = "Theoretical Quantiles",
     y = "Sample Quantiles"
   ) + theme_classic() + ggtitle("Slope + Intercept")+
annotate("text", x = -1, y = 1.5, label = paste0("KS test: p = ",round(ks,3)), color = "black", size = 3)+
annotate("text", x = -1, y = 1, label ="Deviation: p = n.s.", color = "black", size = 3)+
annotate("text", x = 0, y = 0.2, label = paste0("Dispersion test: p = ",round(p_value,3)), color = "black", size = 3)+
annotate("text", x = 0, y = -0.2, label ="Deviation: p = n.s.", color = "black", size = 3)+
annotate("text", x = 1, y = -1, label = paste0("Outlier test: p = 1"), color = "black", size = 3)+
annotate("text", x = 1, y = -1.5, label ="Deviation: p = n.s.", color = "black", size = 3)

```

```{r}
#| echo: false
#| include: false
#| error: false
#| warning: false
resid<-residuals(dlm_aksalmon_2)
resid_dlm <- predict(dlm_aksalmon_2)$pred
resid_dlm$resid <- resid_dlm$y - resid_dlm$estimate
dat$n_date <- as.numeric(as.factor(dat_aksalmon$date))
resid_dlm$date <- dat_aksalmon$date

p<-summary(gam(resid_dlm$resid~s(dat_aksalmon$date)))$s.pv
dlm_resid2<- ggplot(resid_dlm, aes(date, resid)) + 
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x)) + 
  theme_classic() + xlab("Date") + ylab("Residuals") + ggtitle("DLM Residuals")+
  annotate("text", x = 1980, y = -3, label =paste0("Temporal Trend: p = ",round(p,3)),color = "black", size = 3)+
  annotate("text", x = 1980, y = -3, label =paste0("Temporal Trend: p = ",round(p,3)),color = "black", size = 3)

## KS Test
ks<- ks.test(
  x = resid$.std.resids,
  y = "pnorm", 
  mean = mean(resid$.std.resids, na.rm = TRUE), 
  sd = sd(resid$.std.resids, na.rm = TRUE)
)$p.value

# Dispersion test
clean_resid <- na.omit(resid$.std.resids)
n_obs <- length(clean_resid)                  # Number of data points
param_count <- dlm_aksalmon_2$num.params                # Total estimated parameters
df_residual <- n_obs - param_count            # Degrees of freedom
chi_stat <- sum(clean_resid^2)
p_value <- pchisq(chi_stat, df = df_residual, lower.tail = FALSE)
cat("Chi-Square Stat:", chi_stat, "\n",
    "Degrees of Freedom:", df_residual, "\n",
    "P-value for Overdispersion:", p_value, "\n")

# outlier
res_obj <- MARSSresiduals(dlm_aksalmon_2)
obs_res <- res_obj$std.residuals[1, ] # Extract for the first response variable
obs_res <- obs_res[!is.na(obs_res)]   # Remove missing values
n_time <- length(obs_res)
alpha_corrected <- 0.05 / n_time
z_critical <- qnorm(1 - alpha_corrected / 2)
total_outliers <- sum(abs(obs_res) > z_critical)
prob_outlier <- alpha_corrected
binomial_test <- binom.test(
  x = total_outliers,
  n = n_time,
  p = prob_outlier,
  alternative = "greater" # Testing if we have MORE outliers than expected
)

dlm_qq_plot2 <-ggplot(resid, aes(sample = .std.resids)) + 
  stat_qq() + stat_qq_line(col='red') + labs(
     x = "Theoretical Quantiles",
     y = "Sample Quantiles"
   ) + theme_classic() + ggtitle("Slope")+
annotate("text", x = -1, y = 1.5, label = paste0("KS test: p = ",round(ks,3)), color = "black", size = 3)+
annotate("text", x = -1, y = 1, label ="Deviation: p = n.s.", color = "black", size = 3)+
annotate("text", x = 0, y = 0.2, label = paste0("Dispersion test: p = ",round(p_value,3)), color = "black", size = 3)+
annotate("text", x = 0, y = -0.2, label ="Deviation: p = n.s.", color = "black", size = 3)+
annotate("text", x = 1, y = -1, label = paste0("Outlier test: p = 1"), color = "black", size = 3)+
annotate("text", x = 1, y = -1.5, label ="Deviation: p = n.s.", color = "black", size = 3)

```

```{r echo = FALSE}
dlm_qq<-ggplot(resids, aes(sample = resid)) + stat_qq() + stat_qq_line() + labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) + theme_classic() + facet_wrap(~ Model,scale="free_y") + ggtitle("DLM Residuals")
```

{{< pagebreak >}}

## Figures and Tables

::: {#suppfig-TSsalmon}
```{r, echo=FALSE}
salmon_DFA
```

Time series of salmon catch from Lizow et al. 2018
:::

::: {#suppfig-TSsst}
```{r, echo=FALSE}
SST_TS
```

Time series of sea surface temperature anomalies from Lizow et al. 2018
:::

::: {#suppfig-TSslm}
```{r, echo=FALSE}

TS_slm
```

Time series of scaled salmon catch trend (salmon color) and SST anomalies (dark blue)
:::

::: {#suppfig-fitslm}
```{r, echo=FALSE}
Fit_slm
```

The relationship between salmon catch and SST anomalies before and after the 1988/1989 time periods
:::

::: {#suppfig-tvintgam}
```{r, echo=FALSE}
int_panel_gam
```

Time-dependent intercept of the GAM model representing the relationship between SST anomalies and salmon catch.
:::

::: {#suppfig-tvslopegam}
```{r, echo=FALSE}
slope_panel_gam
```

Time-dependent slope of the GAM model representing the relationship between SST anomalies and salmon catch.
:::

::: {#suppfig-gamresid}
```{r, echo=FALSE}
#| fig-height: 6
gamresid
```

Time series of residuals from the GAM model fit with a time-dependent slope and both a time-dependent slope and intercept. p-value denotes the significance of the trend in residuals through time. 
:::
:::

::: {#suppfig-gamqq}
```{r, echo=FALSE}
#| fig-width: 7
#| out-width: "100%"

par(mfrow = c(1, 2))
plotQQunif(simulationOutputcr1,main = "Intercept + slope")
plotQQunif(simulationOutputcr2,main = "Slope")
par(mfrow = c(1, 1))
```

QQ plot from the GAM model fit with a time-dependent slope and both a time-dependent slope and intercept.
:::


::: {#suppfig-lfofold}
```{r, echo=FALSE}
#| fig-width: 6
#| fig-height: 7
ggarrange(tvintsl_lfo,tvsl_lfo,nrow=2,ncol=1,labels=c("A.","B."))

```

Leave-future-out cross validation for the GAM models fit with (A.) both a time-dependent slope and intercept, (B.) a time-dependent slope and and (C.).
:::

::: {#suppfig-tvint}
```{r, echo=FALSE}
int_panel
```

Time-dependent intercept of the DLM model representing the relationship between SST anomalies and salmon catch.
:::

::: {#suppfig-tvslope}
```{r, echo=FALSE}
slope_panel
```

Time-dependent slope of the DLM model representing the relationship between SST anomalies and salmon catch.
:::

::: {#suppfig-dlmresid}
```{r, echo=FALSE}
#| fig-height: 6
dlm_resids
```

Time series of residuals from the DLM model fit with a time-dependent slope and both a time-dependent slope and intercept. p-value denotes the significance of the trend in residuals through time. 
:::
:::

::: {#suppfig-dlmqq}
```{r, echo=FALSE}
ggarrange(dlm_qq_plot1,dlm_qq_plot2)
```

QQ plot from the DLM model fit with a time-dependent slope and both a time-dependent slope and intercept.
:::

{{< pagebreak >}}

::: {#supptab-AIC}
```{r, echo=FALSE}
knitr::kable(head(AIC_tab))
```

AIC values for five linear models
:::


::: {#supptab-results}
```{r, echo=FALSE}

  knitr::kable(mod.results)
```

Summary of the output of the best supported model (model 4)
:::

::: {#supptab-rmse}
```{r, echo=FALSE}
  knitr::kable(cv)
```

RMSE for LFO-CV for both test and training datasets for the time-dependent slope and intercept, and the time-dependent slope models.
:::
