---
title: "Appendix 3: Gulf of Alaska salmon analysis"
format: pdf
editor: source
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
    - kind: float
      key: supptab
      latex-env: supptab
      reference-prefix: Table S
      space-before-numbering: false
      latex-list-of-description: Supplementary Table
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  warning = FALSE,  
  error = FALSE,    
  message = FALSE
)
library(paletteer)
library(MARSS)
library(mgcv)
library(dplyr)
library(forecast)
library(ggplot2)
library(lubridate)
library(marssTMB)
library(tidyverse)
library(strucchange)
library(data.table)
```

Here we show the LM, GAM, and DLM analysis for the Gulf of Alaska salmon dataset. For complete code, see the [full quarto document](https://github.com/ecosystem-state/nonstationarity-best-practices/blob/main/GOA_salmon_case_study/Appendix3_GoASalmon.qmd)

## Data Processing

First, we standardize and examine the trends in SST and salmon abundace from Litzow et al. 2018. 

```{r, echo=FALSE}
# Load salmon timeseries collapsed into DFA trend
salm.dat <- readRDS(here::here("03_GOA_salmon_case_study/Data/dfa_trend.rds")) 

# Load 3-year running mean winter SST timeseries
sst.dat <- readRDS(here::here("03_GOA_salmon_case_study/Data/winterSST_3yr_running_mean.rds")) 

# Bind
right_join(salm.dat, sst.dat) -> dat

# Plot salmon DFA
salmon_DFA<- ggplot(dat, aes(year, salmon_DFA)) +
  theme_bw() +
  geom_line() +
  geom_hline(yintercept = 0) +
  geom_point() +
  geom_errorbar(aes(x=year, ymin=conf.low, ymax=conf.high)) +
  xlab("") +
  ylab("Salmon Trend")

# Plot sst
SST_TS<- ggplot(dat, aes(year, sst_3yr_running_mean)) +
  theme_bw() +
  geom_line() +
  geom_point() +
  xlab("") +
  ylab("3-year running mean SST")

# Scale predictor sst
dat %>%
  mutate(z_sst = scale(sst_3yr_running_mean)) %>%
  na.omit() -> dat2
```

::: {#suppfig-TSsalmon}
```{r, echo=FALSE}
salmon_DFA
```
Time series of salmon catch from Lizow et al. 2018
:::

::: {#suppfig-TSsst}
```{r, echo=FALSE}
SST_TS
```
Time series of sea surface temperature anomalies from Lizow et al. 2018
:::

## LM

We start our investigation using linear models, to evaluate the relationship between salmon catch and SST anomalies  with periods defined a-priori as before and after 1988/1989 (from Litzow et al. 2018). Previous investigations used this approach, and while we primarily discuss DLMs and GAMs for this dataset in the main text, we include this illustrative example here for completeness and comparability to the previous study. 

We compare linear model structures using classic model selection criteria, AIC. First, we examine a 1) intercept only model, 2) salmon abundance with period as a predictor, 3) salmon abundance with period and SST anomalies as predictors, and 4) an interaction between SST anomalies and period. This shows that model 2 is a better fit to the data than model one illustrating that a time-varying intercept is an improvement to model fit than a time-invariant intercept. Similarly, model 4 is the best model showing that a time-varying slope with SST is an improvement to a time-invariant slope (@supptab-AIC).

```{r, echo=FALSE}
dat2 %>%
  mutate(period = case_when((year <= 1988) ~ "early",
                            (year > 1988) ~ "late")) -> dat2

mod_period<-lm(salmon_DFA~as.factor(period), data=dat2) # with period
mod_int<-lm(salmon_DFA~1, data=dat2) # intercept only
mod_sst.period.int<-lm(salmon_DFA~z_sst*as.factor(period), data=dat2) # with interaction
mod_sst.period<-lm(salmon_DFA~z_sst+as.factor(period), data=dat2) # no interaction
mod_sst.period.noint<-lm(salmon_DFA~0+z_sst:as.factor(period), data=dat2) # no interaction


mod_int_AIC <- AIC(mod_int)
mod_mod_period <- AIC(mod_period) 
mod_sst.period_AIC <- AIC(mod_sst.period)
mod_sst.period.int_AIC <- AIC(mod_sst.period.int) 
mod_sst.period.noint_AIC <- AIC(mod_sst.period.noint) 

model_names <- c("1. Intercept only", "2. Intercept by period only \n (time-varying intercept)", 
"3. Intercept by period and SST", "4. Interaction between SST and period \n (time-varying slope and intercept)", "5. Time-varying slope")
AIC <- c(mod_int_AIC, mod_mod_period, mod_sst.period_AIC,mod_sst.period.int_AIC,mod_sst.period.noint_AIC)
AIC_tab <-data.frame(Model = model_names, AIC = round(AIC, 2))

```

::: {#supptab-AIC}

```{r, echo=FALSE}
knitr::kable(head(AIC_tab))
```
AIC values for four linear models
::: 

{{< pagebreak >}}

Functionally, model 4 has a time-varying slope and a time-varying intercept which we can see by both plotting the relationships and examining the model output. We can see a strong positive correlation prior to 1988 (@suppfig-TSslm) followed by a weak negative relationship which is represented in the slopes of the relationship (@supptab-results & @suppfig-fitslm).

```{r, echo = FALSE}

mod.results <-data.table(round(coef(summary(mod_sst.period.int)),2), keep.rownames = 'term')

```

::: {#supptab-results}
```{r, echo=FALSE}

  knitr::kable(mod.results)
```
Summary of the output of the best supported model (model 4)
:::

```{r, echo = FALSE}


Fit_slm<-ggplot(data=dat2, aes(group=period,col=period, y=salmon_DFA,x=z_sst)) +
  stat_smooth(method = "lm")+
  scale_colour_discrete(labels=c("before 1988/89", "after 1988/89"))+
  geom_point()+
  ylab("Salmon trend")+
  xlab("SST")+ theme_bw()

TS_slm<- ggplot()+
  geom_line(dat2, mapping = aes(x=year, y=scale(salmon_DFA)), color = "salmon", size = 1) +
  geom_line(dat2, mapping = aes(x = year, y = z_sst), color = "darkblue", size = 1)+
  geom_point(dat2, mapping = aes(x=year, y=scale(salmon_DFA)), color = "salmon") +
  geom_point(dat2, mapping = aes(x = year, y = z_sst), color = "darkblue")+
  scale_colour_discrete(labels=c("salmon trend", "SST"))+
  geom_vline(xintercept=1988, lty=2)+
  xlab("Year")+ theme_bw()

```

::: {#suppfig-TSslm}
```{r, echo=FALSE}

TS_slm
```
Time series of scaled salmon catch trend (salmon color) and SST anomalies (dark blue)
:::

::: {#suppfig-fitslm}
```{r, echo=FALSE}
Fit_slm
```
The relationship between salmon catch and SST anomalies before and after the 1988/1989 time periods
:::

{{< pagebreak >}}

## GAM

We fit two distinct GAM models to the salmon catch data and SST anomalies to illustrate the differences in a model with a time-varying slope and a time-varying intercept, and a model that only has a time-varying slope. We find that when fitting a time-varying slope and time-varying intercept model, most of the variability in the relationship is explained by the time-varying intercept (@suppfig-tvintgam) while the time-varying slope does not change much through time (@suppfig-tvslopegam). When only a time-varying slope is fit to the data, the slope parameter is much more variable through time (@suppfig-tvslopegam), and looks similar to the slope estimated by the linear models (@suppfig-fitslm). 

```{r,include=FALSE}
dat_aksalmon <- readRDS("data_for_AK_salmon_example.rds")
col<-paletteer_d("nationalparkcolors::Everglades")
#renaming
dat_aksalmon <- dplyr::rename(dat_aksalmon,
                              response = salmon_DFA,
                              driver = z_sst, 
                              date = year)
K_AK <- round(0.99*nrow(dat_aksalmon))
dat_aksalmon$n_date <- seq(1,nrow(dat_aksalmon))
gam_aksalmon_1 <- gam(response ~ s(n_date,k=K_AK) + s(n_date, by = driver, bs = "gp",k=K_AK), data = dat_aksalmon)
gam_aksalmon_2 <- gam(response ~ s(n_date, by = driver, bs = "gp",k=K_AK), data = dat_aksalmon)

# extract the smooth estimates with SEs to compare to DLM
plot_3 <- plot(gam_aksalmon_1, seWithMean = TRUE, n = nrow(dat_aksalmon))
plot_4 <- plot(gam_aksalmon_2, seWithMean = TRUE, n = nrow(dat_aksalmon))

coef_gam_AK1 <- data.frame(date = dat_aksalmon$date,
                        int_est = plot_3[[1]]$fit,
                        int_se = plot_3[[1]]$fit,
                        slope_est = plot_3[[2]]$fit,
                        slope_se = plot_3[[2]]$fit,
                        time_varying = "Intercept + slope",
                        dataset = "Alaska salmon : GAM")
coef_gam_AK2 <- data.frame(date = dat_aksalmon$date,
                        int_est = NA,
                        int_se = NA,
                        slope_est = plot_4[[1]]$fit,
                        slope_se = plot_4[[1]]$se,
                        time_varying = "Slope",
                        dataset = "Alaska salmon : GAM")

coefs_gam <- rbind(coef_gam_AK1, coef_gam_AK2)

slope_panel_gam <-ggplot(coefs_gam, aes(date, slope_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=slope_est-2*slope_se, ymax = slope_est+2*slope_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-varying slope") + 
  xlab("") + theme_bw() + 
  ggtitle("Slope")+
  #ylim(c(-5,11))+
  facet_wrap(~ dataset, scale="free") + 
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  #guides(fill="none",colour="none",lty="none")+
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )

int_panel_gam <- ggplot(coefs_gam, aes(date, int_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=int_est-2*int_se, ymax = int_est+2*int_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-varying intercept") + 
  xlab("") + theme_bw() + 
  facet_wrap(~ dataset, scale="free") + 
  ggtitle("Intercept")+
  #ylim(-20,10.5)+
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )


```

::: {#suppfig-tvintgam}
```{r, echo=FALSE}
int_panel_gam
```
Time-varying intercept of the GAM model representing the relationship between SST anomalies and salmon catch. 
:::


::: {#suppfig-tvslopegam}
```{r, echo=FALSE}
slope_panel_gam
```
Time-varying slope of the GAM model representing the relationship between SST anomalies and salmon catch. 
:::



```{r,include=FALSE}


dat <- dat_aksalmon
m <- 2
TT <- nrow(dat)
B <- diag(m)  ## 2x2; Identity
U <- matrix(0, nrow = m, ncol = 1)  ## 2x1; both elements = 0
Q <- matrix(list(0), m, m)  ## 2x2; all 0 for now
diag(Q)[1] <- c("q.alpha")
diag(Q)[2] <- c("q.beta")
Z <- array(NA, c(1, m, TT))  ## NxMxT; empty for now
Z[1, 1, ] <- rep(1, TT)  ## Nx1; 1's for intercept
Z[1, 2, ] <- dat$driver  ## Nx1; predictor variable
A <- matrix(0)  ## 1x1; scalar = 0
R <- matrix("r")  ## 1x1; scalar = r
## only need starting values for regr parameters
inits_list <- list(x0 = matrix(c(0, 0), nrow = m))
## list of model matrices & vectors
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
# convert response to matrix
dat_mat <- matrix(dat$response, nrow = 1)
# fit the model -- crank up the maxit to ensure convergence
dlm_aksalmon_1 <- MARSS(dat_mat, inits = inits_list, model = mod_list,
               control = list(maxit=4000), method="TMB")

# Fit a second model with just time-varying slope
diag(Q)[1] <- 1e-10
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
dlm_aksalmon_2 <- MARSS(dat_mat, inits = inits_list, model = mod_list,
               control = list(maxit=4000), method="TMB")

coef_dlm_AK1 <- data.frame(date = dat_aksalmon$date,
                        int_est = dlm_aksalmon_1$states[1,],
                        int_se = dlm_aksalmon_1$states.se[1,],
                        slope_est = dlm_aksalmon_1$states[2,],
                        slope_se = dlm_aksalmon_1$states.se[2,],
                        time_varying = "Intercept + slope",
                        dataset = "Alaska salmon : DLM")
coef_dlm_AK2 <- data.frame(date = dat_aksalmon$date,
                        int_est = NA,
                        int_se = NA,
                        slope_est = dlm_aksalmon_2$states[2,],
                        slope_se = dlm_aksalmon_2$states.se[2,],
                        time_varying = "Slope",
                        dataset = "Alaska salmon : DLM")

coefs <- rbind(coef_dlm_AK1, coef_dlm_AK2)

slope_panel <-ggplot(coefs, aes(date, slope_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=slope_est-2*slope_se, ymax = slope_est+2*slope_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-varying slope") + 
  xlab("") + theme_bw() + 
  ggtitle("Slope")+
  #ylim(c(-5,11))+
  facet_wrap(~ dataset, scale="free") + 
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
 # guides(fill="none",colour="none",lty="none")+
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )

int_panel <- ggplot(coefs, aes(date, int_est, group = time_varying, fill=time_varying, col=time_varying)) + 
  geom_ribbon(aes(ymin=int_est-2*int_se, ymax = int_est+2*int_se), alpha=0.3, col = NA) + 
  geom_line(aes(lty=time_varying)) + 
  ylab("Time-varying intercept") + 
  xlab("") + theme_bw() + 
  facet_wrap(~ dataset, scale="free") + 
  ggtitle("Intercept")+
  #ylim(-20,10.5)+
  #scale_color_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  #scale_fill_viridis_d(option="magma",begin=0.2, end=0.8, name = "Time-varying") + 
  scale_fill_manual(values=c(col[4],col[5]))+
  scale_colour_manual(values=c(col[4],col[5]))+
  
  theme(
    strip.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )


```


## DLM

We fit two distinct DLM models to the salmon catch data and SST anomalies to compare to the results of the GAM models. Generally, the parameters estimated by the DLM are more certain and follow similar patterns through time as the GAM (@suppfig-tvint & @suppfig-tvslope). There is also more interannual variability in the parameter estimates of the DLM compared to the continuously smoothed pattern of the GAM. 

::: {#suppfig-tvint}
```{r, echo=FALSE}
int_panel
```
Time-varying intercept of the DLM model representing the relationship between SST anomalies and salmon catch. 
:::

::: {#suppfig-tvslope}
```{r, echo=FALSE}
slope_panel
```
Time-varying slope of the DLM model representing the relationship between SST anomalies and salmon catch. 
:::
